### 2015-03-29 markdown/manuscript.tex
third pass, intro

More than a decade ago, Rick Riolo, Bill Worzel and I were working on a consulting project together that involved evolutionary algorithms and genetic programming. As we were chatting one day, Bill asked Rick what he'd most like to see as part of the research program of GP ``moving forward''. Rick's answer informs this contribution, as well as much of my professional work with GP in the years since.

As I recall, Rick answered that he'd like to better understand the ``symptoms'' we often see when we run an evolutionary search process: premature convergence, failure to improve, catastrophic lack of diversity, and the sense when we look at results or ongoing runs that some particular choices we've made setting the many parameters \emph{just aren't quite right}. There are numerous well-written papers suggesting ways around local minima, proposing reasonable and exotic search operators, and running ``horse races'' between variant algorithms on benchmark problems. But years later, I take Rick's challenge not only to mean that it might be useful to have a benchmarking catalog which lists conditions in which search operator $X$ acting under contingency $Y$ tends to produce outcome $Z$, but also that \emph{the things we identify as symptoms themselves} are poorly understood, to the point that we think of them as ``subjective''.

Somewhere during that same project, I remember a Project Manager telling the story that when domain expert customers were shown a collection of Pareto-optimal solutions to the problem being explored, they were confused. ``We don't want a choice, we want \emph{the best}.'' This came after months of analysis, interviews, and a collective agreement that the problem was fundamentally multi-objective. But when the decision-makers were faced with the unquestionably successful results they had collectively specified, \emph{those results weren't right enough}.

In this contribution we must face the fact that such a response is inevitable: not just from silly lay ``customers'' in an application project, but from ourselves; whether the project is ``theoretical'' or ``practical''; whether it is ``small'' or ``large''. Projects resist our initial plans.

The ``field'' of GP has grown quite a bit in the intervening 15 years since these two anecdotes. All of us who follow it closely have played a role in the expanding front of new methodologies and techniques---the extraordinary ones we describe in our little workshop, and also the newsworthy and admirable public successes.

But this acceleration of practical successes has been accompanied by a dilution of the scope of the theoretical warrants we use to justify them: Most examples of ``practical success'' involve unique domain-specific quirks, each implementation makes numerous contingent (and sometimes arbitrary) design and architectural choices, and even the ontological foundations of ``individual'', ``fitness'', ``behavior'' and ``population'' start to get sloppy around the edges whenever we actually \emph{look} at What People Actually Do to Solve Problems With GP.

While this can be disappointing for those of us who follow the progress of theoretical work closely, it has more serious consequences. The lack of interest in GP among statisticians, planners, designers and mathematical programming aficionados is, I argue, not just a matter of disciplinary border wars. I suspect rather that skepticism arises from our own inability to say what's \emph{happening} in any particular GP run---\emph{even when we look}. Nor for that matter can we be very helpful in explaining \emph{why}, or what to do in response to any given contingency. We have learned that \emph{more CPU time} can sometimes get us past a tricky stage of ``not working'', but any project that lives very far from the bounds of the last century's version of ``genetic programming'' is unexplored territory.

This is not the fault of GP as a field, nor of our theoreticians and engineers. Rather I'll argue that a deep philosophical misconception presides over the work, both within and outside our field. This misconception undermines our understanding of what GP is \emph{doing}, of the way it unfolds in theory and in practice, and even what it's \emph{for}.

I will make my case here in the form of an exercise, or \emph{kata}. This is not intended to be a ``thought-experiment'', nor is it a suggestion for a new way of working. Rather it is a formal exercise to be undertaken by those of us already working closely with GP systems. In the immediate case, my intent is to surface the three problems I've identified above:

\begin{enumerate}
\item What GP \emph{does}
\item How GP \emph{unfolds}
\item What GP is \emph{for}
\end{enumerate}

The rules for this exercise, ``Doing GP as if you meant it'', will feel the most artificial and restrictive to those of us with the most experience. Like the martial arts exercises from which it is ultimately inspired, it isn't intended to be simple or even pleasant for the participants. But because it emphasizes the interaction between the GP system and the user, it not only surfaces the philosophical mistake I've mentioned, but immediately suggests tools with which we can address Rick's decade-old wish.

>>>>

More than a decade ago, Rick Riolo, Bill Worzel and I were working on a consulting project together that involved genetic programming. As we chatted one day, Rick was asked what he'd most like to see as part of the research program of GP ``moving forward''. Rick's answer informs this contribution, as well as much of my professional work with GP in the years since.

As I recall, Rick said he'd like work to focus on the ``symptoms'' we often see in evolutionary search processes: premature convergence, failure to improve, catastrophic lack of diversity, mysterious things like that sense we all get when we look at results that suggests \emph{we made bad choices} of some sort.

The literature abounds with well-written papers describing tips for avoiding local minima, improving on common search operators, and describing ``horse races'' between Bad Old and Better New search methodologies applied to benchmark problems. But I take Rick's challenge not only to mean that it would be useful to have a catalog which lists situations where search operator $X$ acting under contingency $Y$ tends to produce outcome $Z$, but also that \emph{the things we identify as symptoms themselves} are poorly understood.

This contribution aims squarely at that sensibility which sees the ``symptoms'' and ``pathologies'' as something we as a research community should be trying to stamp out \emph{before} letting users run their own GP systems. Instead, I'll argue that because GP differs qualitatively and philosophically from most other machine learning approaches, and also from the broader lay understanding of the ``automated design'' and exploratory research, much of the value GP offers is lost or ignored when we approach it as something to be ``cleaned up''.

\subsection{Challenges}\hypertarget{challenges}{}\label{challenges}

Somewhere during that same project years ago, I remember a Project Manager recounting the story that when domain expert customers were shown a collection of Pareto-optimal solutions to the problem being explored, they were upset and confused: ``We don't want a choice, we want \emph{the best}.'' Of course there could be no \emph{best}, and they supposedly knew that; this response came after \emph{months} of analysis, interviews, technical planning and a collective agreement that the problem was fundamentally multi-objective. Even so, when the decision-makers were faced with the \emph{unquestionably} successful results they had collectively specified and specifically asked for\ldots{} those results weren't right enough.

I would much rather we faced the fact that this shocked response to \emph{seeing what's happening} is inevitable---not just in the form of unwitting surprise from silly lay ``customers'' in an application project, but also from ourselves when we undertake any \emph{interesting} GP project, whether the project is ``theoretical'' or ``practical'', ``small'' or ``large''.

Any \emph{interesting} project will resist our best expectations. Now and then we should question whether our habitual response to that inevitable resistance is in any sense effective.

The field of GP has grown quite a bit in the 15 years since my two anecdotes. We who follow it closely have watched as frameworks, techniques and methodologies have diversified. There are the extraordinary ones we describe to one another in our little workshops, but also an increasing number of newsworthy and admirable public successes.

But this acceleration of innovation brings with it a dilution of the theoretical warrants we use to justify results. Most practical successes today highlight unique domain-specific quirks. New language and framework implementations makes numerous contingent (and often arbitrary) design and architectural choices. Even the fundamental ontological concepts of ``individual'', ``fitness'', ``behavior'', ``generation'' and ``population'' start to get troubling when try to fit many modern algorithms into their categories.

This growing divergence between current theory and practice has consequences both outside and within our field. The continued lack of interest in GP among statisticians, planners, designers and fans of traditional mathematical programming models is not merely a matter of disciplinary border wars. As we increasingly lose the ability to say \emph{why} something is happening in a given GP project, we are faced with an audience who stopped paying attention to progress in GP last century. As a result, students and researchers aiming to convince peers outside the field are forced to undertake mismatched experiments and apply unconvincing statistical tests. In response to these constant distractions, much time and attention goes towards unhelpful work guaranteeing that which cannot be guaranteed: running ``replicates'' of a process that is \emph{designed} to provide novel answers, measuring ``reliability'' of a process that \emph{intentionally} skirts dynamical chaos, providing ``summary charts'' of a process that strives to be as complex as the evolution of living beings, and (possibly worst) setting strict deadlines for an open-ended process to ``finish'' or ``succeed''.

I hasten to say this is not a fault in our field, but rather a broader philosophical and cultural problem. But it is nonetheless a misconception that undermines our understanding of what GP is \emph{doing}, of the way it unfolds in theory and in practice, and even what it's \emph{for}.

\subsection{A top-down approach}\hypertarget{a-top-down-approach}{}\label{a-top-down-approach}

I will make my case here in the form of an exercise for advanced GP users, or \emph{kata}. The habit of pursuing \emph{kata}, ``code retreats'', ``hackathons'' and other skill-honing practices is popular among software developers, and especially among the more advanced. Indeed, the title of my exercise (``GP as if you meant it'') is taken from a particularly influential exercise designed by Keith Braithwaite, ``TDD as if you meant it''.

Braithwaite's exercise feels subjectively \emph{harder} for more advanced programmers honing their development form; he suggests that novice programmers haven't learned ingrained but questionable habits, and haven't identified ``shortcuts'' that ``simplify'' the practices. In the same way, this exercise will feel \emph{most} artificial and restrictive to those of us with the most experience with GP. But like the martial arts exercises from software \emph{kata} were inspired, it isn't intended to be simple or even pleasant for the participants.

I intend it to expose habits I'll tactlessly call ``psueudo-GP'' among those of us who have learned through the years to think it's cheap and painless to \emph{just shut it off and start over} when things start to get strange in the course of a GP project. As an important side-effect, it helps surface that philosophical problem I alluded to above. But I have found the most useful result is how strongly and immediately it suggests tools with which one can address Rick's decade-old wish. It forces the participant to formally identify ``pathology'' and ``symptom'' before allowing them to attempt a ``fix''\ldots{} and even that must done with an intentionally limited set of tools.

Note, what I'll describe is a ``thought-experiment'', nor is it a serious suggestion for a new way of working on ``real problems'', nor as ``training'' for newcomers to the field. Rather it is designed as a rigorous and formal exercise to be undertaken by those of us already working closely with GP systems. Its intent is to surface the three shortcomings I've identified above:

\begin{enumerate}
\item what GP \emph{does}
\item how GP \emph{unfolds}
\item what GP is \emph{for}
\end{enumerate}
